---
title: "COVID-19 Data"
author: "Vaughn Pourchot"
date: "2023-10-16"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=TRUE}
library(tidyverse)
library(tidyr)
library(dplyr)
library(stringr)
library(lubridate)
url_in <- "https://raw.githubusercontent.com/CSSEGISandData/COVID-19/master/csse_covid_19_data/csse_covid_19_time_series/"
file_names <- c("time_series_covid19_confirmed_US.csv",
                "time_series_covid19_confirmed_global.csv",
                "time_series_covid19_deaths_US.csv",
                "time_series_covid19_deaths_global.csv")
urls <- str_c(url_in,file_names)
```

## Import Data

``` {r import_data, message = FALSE, include=TRUE}
global_cases <- read.csv(urls[2])
global_deaths <- read.csv(urls[4])
US_cases <- read.csv(urls[1])
US_deaths <- read.csv(urls[3])
```
## R Markdown

```{r tidy_global_data, include=TRUE}
global_cases <- global_cases %>%
  pivot_longer(cols = starts_with("X"),
               names_to = "date",
               values_to = "cases")
global_cases$date <- gsub("^X", "", global_cases$date)
global_cases$date <- as.Date(gsub("\\.", "-", global_cases$date), format = "%m-%d-%y")
```

```{r tidy_global_deaths, include=TRUE}
global_deaths <- global_deaths %>%
  pivot_longer(cols = starts_with("X"),
               names_to = "date",
               values_to = "cases")
global_deaths$date <- gsub("^X", "", global_deaths$date)
global_deaths$date <- as.Date(gsub("\\.", "-", global_deaths$date), format = "%m-%d-%y")
```


```{r tidy_US_deaths, include=TRUE}
US_deaths <- US_deaths %>%
  pivot_longer(cols = starts_with("X"),
               names_to = "date",
               values_to = "deaths")
US_deaths$date <- gsub("^X", "", US_deaths$date)
US_deaths$date <- as.Date(gsub("\\.", "-", US_deaths$date), format = "%m-%d-%y")


```

```{r remove_columns, include=TRUE}
US_deaths <- US_deaths %>%
  select(-UID,-iso2,-iso3,-code3,-FIPS,-Lat,-Long_,-Combined_Key)
```

```{r tidy_US_cases, include=TRUE}
US_cases <- US_cases %>%
  pivot_longer(cols = starts_with("X"),
               names_to = "date",
               values_to = "cases")
US_cases$date <- gsub("^X", "", US_cases$date)
US_cases$date <- as.Date(gsub("\\.", "-", US_cases$date), format = "%m-%d-%y")
US_cases <- US_cases %>%
  select(-UID,-iso2,-iso3,-code3,-FIPS,-Lat,-Long_,-Combined_Key)
```


```{r join_global_cases_deaths, include=TRUE}
global <- global_cases %>%
  full_join(global_deaths, by = c("Province.State", "Country.Region", "Lat", "Long", "date")) %>%
  rename(Country_Region = 'Country.Region', 
         Province_State = 'Province.State') %>%
  mutate(date = ymd(date))
```

```{r filter_out_empty, include=TRUE}
global <- global %>% 
  rename(cases = cases.x) %>%
  rename(deaths = cases.y) %>%
  filter(cases >0) %>%
  select(-Lat,-Long)
```

```{r US_combine, include=TRUE}
US <- US_cases %>%
 full_join(US_deaths)

```

clean up x's
```{r clean_up_x, include=TRUE}
global_cases$date <- gsub("^X", "", global_cases$date)
global_cases$date <- as.Date(gsub("\\.", "-", global_cases$date), format = "%m-%d-%y")
```


```{r tidy_global_data_combined, include=TRUE}
global_cases <- global_cases %>%
  mutate(date = as.character(date), cases = as.character(cases)) %>% # Coerce to the same type
  pivot_longer(
    cols = -c('Province.State', 'Country.Region', Lat, Long),
    names_to = "date",
    values_to = "cases"
  ) %>%
  select(-c(Lat, Long))

```

```{r combined_key_global, include=TRUE}
global <- global %>%
  unite("Combined_Key",
        c(Province_State, Country_Region),
        sep = ", ",
        na.rm = TRUE,
        remove = FALSE)

```


```{r add_pop_data_global, include=TRUE}
uid_lookup_url <- "https://github.com/CSSEGISandData/COVID-19/raw/master/csse_covid_19_data/UID_ISO_FIPS_LookUp_Table.csv"
uid <- read.csv(uid_lookup_url) %>%
 select(-c(Lat,Long_,Combined_Key,code3,iso2,iso3,Admin2))
global <- global %>%
 left_join(uid, by = c("Province_State","Country_Region")) %>%
 select(-c(UID,FIPS)) %>%
 select(Province_State,Country_Region,date,cases,deaths,Population,Combined_Key)
```


```{r visualization_grouping_by_state, include=TRUE}
US_by_state <- US %>%
  group_by(Province_State, Country_Region, date) %>%
  summarize(cases = sum(cases), 
            deaths = sum(deaths), 
            Population = sum(Population),
            .groups = "drop") %>%
  mutate(deaths_per_mil = deaths * 1000000 / Population) %>%
  select(Province_State, Country_Region, date, cases, deaths, deaths_per_mil, Population)


```


```{r visualization_grouping_total_US, include=TRUE}
US_totals <- US_by_state %>%
  group_by(Country_Region, date) %>%
  summarize(cases = sum(cases), 
            deaths = sum(deaths), 
            Population = sum(Population),
            .groups = "drop") %>%
  mutate(deaths_per_mil = deaths * 1000000 / Population) %>%
  select(Country_Region, date, cases, deaths, deaths_per_mil, Population)


```

```{r visualization_by_total_combined, include=TRUE}
US_totals %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y= deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y=NULL)


```

```{r visualization, include=TRUE}
state <- "New York"
US_by_state %>%
  filter(Province_State == state) %>%
  filter(cases > 0) %>%
  ggplot(aes(x = date, y = cases)) +
  geom_line(aes(color = "cases")) +
  geom_point(aes(color = "cases")) +
  geom_line(aes(y = deaths, color = "deaths")) +
  geom_point(aes(y= deaths, color = "deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y=NULL)


```

```{r add_variables, include=TRUE}
US_by_state <- US_by_state %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
US_totals <- US_totals %>%
  mutate(new_cases = cases - lag(cases),
         new_deaths = deaths - lag(deaths))
```

```{r visualization_by_total, include=TRUE}
US_totals %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y= new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = "COVID19 in US", y=NULL)
```


```{r visualization2, include=TRUE}
state <- "New York"
US_by_state %>%
  filter(Province_State == state) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y= new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y=NULL)


```

```{r visualization_grouping, include=TRUE}
US_state_totals <- US_by_state %>%
  group_by(Province_State) %>%
  summarize(cases = max(cases), 
            deaths = max(deaths), 
            Population = max(Population),
            cases_per_thou = 1000 * cases / Population,
            deaths_per_thou = 1000 * deaths / Population) %>%
  filter(cases > 0, Population > 0)
US_state_totals %>% 
 slice_min(deaths_per_thou,n = 10)  
US_state_totals %>%
 slice_min(deaths_per_thou, n = 10) %>%
 select(deaths_per_thou, cases_per_thou, everything())
US_state_totals %>%
 slice_max(deaths_per_thou, n = 10) %>%
 select(deaths_per_thou, cases_per_thou, everything())
```  

```{r modeling, include=TRUE}
mod <- lm(deaths_per_thou ~  cases_per_thou, data = US_state_totals)

US_state_totals %>% slice_max(cases_per_thou)
x_grid <- seq(1,151)
new_df <- tibble(cases_per_thou = x_grid)
US_state_totals %>% mutate(pred = predict(mod))
US_tot_w_pred <- US_state_totals %>% mutate(pred = predict(mod))
US_tot_w_pred
```

```{r visualization_model, include=TRUE}
US_tot_w_pred %>% ggplot() +
  geom_point(aes(x = cases_per_thou, y = deaths_per_thou), color = "blue") +
  geom_point(aes(x = cases_per_thou, y = pred), color = "red")

```

## Unique Analysis: Maine COVID data

```{r visualization_maine, include=TRUE}
state <- "Maine"
US_by_state %>%
  filter(Province_State == state) %>%
  ggplot(aes(x = date, y = new_cases)) +
  geom_line(aes(color = "new_cases")) +
  geom_point(aes(color = "new_cases")) +
  geom_line(aes(y = new_deaths, color = "new_deaths")) +
  geom_point(aes(y= new_deaths, color = "new_deaths")) +
  scale_y_log10() +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90)) +
  labs(title = str_c("COVID19 in ", state), y=NULL)


```
I live in the state of Maine so I was excited to view the local data and see how we're looking.  It's a little disappointing to see case rates climbing but that was expected.  It's very encouraging to see a distinct trend line with a downward slope for the deaths.  Even with the spike in cases, the deaths appear to be within a reasonable margin of the trend line.